### 2017-05-12 paper/paper.tex
Finalizing before opposition

\begin{abstract}
	The abstract text goes here.
\end{abstract}

@@

\section{Introduction}
\subsection{Background}
Taking a collective decision as a population is difficult. To solve this issue, voting systems with defined rules are used. They are used to show common preferences within a population, for example what politician a population wants to see elected. Several types of systems have been designed and there are a myriad of variations of those systems. They range from simple methods such as “most votes win” to complex processes that can only be practically carried out by a computer. However, practically all voting systems are algorithmic in nature, which makes them interesting to study for both Computer Scientists and Mathematicians. When deciding on what systems that shall be used, it is useful to know what differences and similarities they have.
\subsection{Purpose}
The purpose of this essay is to evaluate differences and similarities in terms of election results in three different voting systems: Single Transferable Vote (STV), First-Past-The-Post (FPTP) and the Schulze Method. This paper will only be examining and discussing differences and similarities in the results of the methods, not touching on how practically feasible the methods might be in actual election.
\pagebreak
\section{Theory}
One would think that by testing all different voting systems and by setting up good criteria for how to evaluate them, it would be possible to come to a definitive conclusion of what system should be used. However, this is not the case. By setting up different mathematical criteria for how a voting system shall behave, it is impossible to satisfy all criteria simultaneously. This is what several theorems, most notably Arrow's Impossibility theorem outlined in 1950, show \autocite{arrow1950difficulty}. This means that any decision regarding choice of electoral system is a balance of their strengths and weaknesses. Simply put: there is no perfect voting system and never will be. This does however not mean that all voting systems are equally efficient, elegant or accurate.

There are several ways of evaluating voting methods that are used. One is the mathematical approach of evaluating what criteria are important to meet and finding elegant methods that satisfy them. Another is by testing and finding probabilities that some criteria are broken — A method that fails to meet a certain criteria only in very few cases is perhaps not that big of a problem. A third way of evaluating methods is not by mathematical criteria, but other metrics such as voter satisfaction or how likely someone is to vote tactically.

@@

All voting methods have been implemented in Typescript – mostly due to the author having previous experience with JavaScript. Typescript is a superset of JavaScript with multiple extra features. Most importantly Typescript has optional static type checking. Sample code in typescript can be seen in code example~\ref{lst:typescript example}.

@@

In order to test the methods and their implementations, test data is needed. There are cases where extensive election data is published, such as in Maltese elections, but full ballot data, which is needed for the implementations has not been found. Instead, the implementations will be tested with data generated from a simple computer algorithm. The algorithm used in this paper is highly primitive, as modeling preferences within a population is far beyond this paper. The full ballot generation program can be found in the \Code{/generator/} folder. The goal of the program is to produce ballots that rank every candidate in order of preference (see figure~\ref{fig:stv ballot} on page~\pageref{fig:stv ballot}). The program represents each voter in a population as belonging to a certain ideology. Each ideology has has a list of weighted preferences indicating how a voter for this ideology is likely to vote. The weighed preferences are generated via an abstraction. A single configuration object is used to create all ideologies.

@@

	\item Based on the base size of each candidate, a probability-list of voting for each candidate is created for every ideology. An ideology's probability-list represents how a voter aligned the ideology is expected to vote.
	\item It creates a certain number of individual ballots. Each ballot is given an ideology and creates its own list preferences based via weighted random number generation according to the probability-list of the assigned ideology.

@@

The output of the program is a set of ballots, where each ballot is an ordered set of preferences. The implementation is highly arbitrary and all constants in the program even more so. However, expanding on the generator program would be beyond the scope of this project. The program works in that it is possible to control variables to test for different election scenarios.

@@

Constructing test scenarios is difficult task. Due to there being so many variables both in real elections and the simulated elections in the generator program, it makes it practically impossible to control all variables. Instead, 3 arbitrary test scenarios will be used. The election will be simulated 1000 times for each scenario and the election data will then run through the different voting methods and the results will be compared. The simulation scenarios are arbitrary but should allow for comparison of the different voting methods. All simulations will have 3 ideologies and 500 voters.

@@

The scenario is built around electing one candidate out of eight. The spread of votes within ideologies is flat and all ideologies are ``weak''. This means that the general spread of votes will be even. An example of the distribution of first-preference votes generated in this scenario can be seen in figure~\ref{fig:example of scenario 1}. The full configuration object can be seen in code example~\ref{lst:scenario 1}.

@@

	\caption{Example distribution of first-preference votes in scenario 1}

@@

	\caption{Example distribution of first-preference votes in scenario 2}

@@

	\caption{Example distribution of first-preference votes in scenario 3}

@@

Three metrics will be used for comparing the results obtained in the different scenarios. The results obtained with each voting method will be compared against the other methods. Comparisons to the configuration files outlined in Sections~\ref{sec:method-scenario1} -~\ref{sec:method-scenario3}.

@@

The standard deviation of the outcomes represents the variance in outcome due to the pseudorandom nature of the generator program. The standard deviation is calculated according to equation~\ref{eq:standard-deviation}. 

@@

	\item Compare the actual outcomes of a certain method with the theoretical result obtained in Step \ref{enum:theoretical}.

@@

The first-past-the-post voting method is a method designed for electing one or several candidates out of a set of candidates. Each voter has one (1) vote which may be allocated to any candidate. In a first-past-the-post election the N candidates with most votes get elected.

@@

The first-past-the-post program is implemented in a single file. The inputs of the program can be seen in code example~\ref{lst:fptp inputs}.

@@

The Single Transferable Vote method is a more complicated algorithm than the first-past-the-post method. Voters are given ballots where they are supposed to rank candidates in order. The specific rules for if you need to rank all candidates or rank several candidates at the same value not can vary. For the sake of simplicity, in this implementation, all candidates must be ranked at unique and linear values. An example of a ballot can be seen in figure~\ref{fig:stv ballot}.	This means that there is more data available for determining the result of the election which the method can take advantage of. Votes are, if needed, transferred between choices within a ballot, hence the name Single Transferable Vote. The method relies on the concept of a quota, the number of votes you need to be elected. The Droop-Quota, defined as $(\frac{\text{total valid poll}}{\text{seats} + 1})+1$, is often used. If a candidate receives equal or more votes than the quota, he/she is elected. If the number of votes exceed the quota, a fraction of the votes are transferred to the next choice on the ballots that voted for the elected candidate. This process is repeated until there are no candidates with more votes than the quota. If there still are seats yet to be filled, the candidate with fewest votes is eliminated and his/her votes are transferred to the next choice on those ballots. This process is repeated until all seats are filled. The process can become quite complex with many ballots and transferring fractional votes. A computer is therefore essential in order to resolve large scale elections.

@@

@@

\tab eliminate $votes_{k}$ and distribute all next-preference votes for candidate $k$ \tab into $votes$\\

@@

The single transferable vote algorithm uses a tree structure to represent the state of the election. Every candidate is a top-level node with an attribute showing the number of current votes for the candidate. Each node has child nodes representing the next preferences of the voters for that particular candidate, see figure~\ref{fig:tree structure}. It is easy to manipulate branches via recursion. Merging and multiplying branches is used when transferring votes from one candidate node to another. A candidate node is a Typescript class. It can be found in the \Code{stv/candidate.ts} file and has the following properties and functions:

@@

The Schulze Method is a method used mainly to determine results in a single-winner election but can also be used to provide rankings and elect multiple candidates in a single election \autocite{Schulze2011}. Ballots are identical to those of the single transferable vote, see figure~\ref{fig:stv ballot} on page~\pageref{fig:stv ballot}. The method is compliant with the Condorcet Criterion, which means that if there is a candidate who the majority prefers in a pairwise comparison with every other candidate, that candidate wins. As mentioned, this method relies on comparing the preferences of every candidate to one another as shown in table~\ref{tab:pairwise comparison matrix}. If there is a Condorcet winner the process is simple: declare the Condorcet winner a winner and run another iteration without that candidate. Repeat this process until all seats are filled. However, there can be Condorcet ties, where there is no Condorcet winner. There are multiple ways of resolving the tie, one of them being the Schulze Method.

@@

The difficult problem in this method is finding the strongest path between candidates. The problem is called the widest path problem in graph theory. An efficient and relatively simple way to compute this problem is via the Floyd-Warshall algorithm which is used in the implementation. See~\ref{lst:schulze algorithm} for the full algorithm.

@@

@@

The Schulze Method is implemented within the file \Code{schulze/index.ts}. It takes two inputs: a list of ballots and the number of seats to be elected. A ballot is an ordered list of candidates. The output of the program is a list of winners. The program first creates the matrices for both the preferences and paths between candidates.
\codeblock{code/schulze/maps.ts}{lst:schulze maps}{Data representation}

@@

Scenario 1 featured 8 candidates where one was to be elected. Vote spread was flat, meaning that votes tend to be diverse in their preferences. This means that there is not an obvious winner present. This can be seen from the first-preference votes in the scenario seen in figure~\ref{fig:example of scenario 1}. The full result in this scenario can be seen in figure~\ref{fig:scenario 1 results}. The outlier in this scenario seems to be the fptp method, electing candidate 0 more often than both other methods. Key metrics for the results in this scenario can be found in table~\ref{tab:scenario 1 result}. Although the misrepresentation metric is not entirely accurate, especially when few seats are elected, it shows that fptp is not entirely representative. The STV and Schulze method seem to return very similar results.

@@

\begin{table}
\centering
\caption{Key metrics for result in scenario 1}
\label{tab:scenario 1 result}
\begin{tabular}{@{}lccc@{}}
\toprule
Method & Standard deviation & Misrepresentation & Most likely winners \\ \midrule
FPTP & 223 & 66 & 2 \\
STV & 315 & 8 & 2 \\
Schulze & 313 & 9 & 2 \\ \bottomrule
\end{tabular}
\end{table}
% [ 223.39259164082, 315.44928276983, 313.6271671906]
\subsection{Scenario 2}
Scenario 2 featured 9 candidates of which 3 were to be elected. The distribution of votes was ``normal'', and as seen in figure~\ref{fig:example of scenario 2}, the election is very close with many candidates having a similar percentage of first-preference votes. The results obtained in scenario 2 are very interesting. FPTP tended to underrepresent candidate 1, and elected a candidate from ideology 2 only in 28\% of elections, despite the ideology theoretically having 30\% of popular support. The Schulze method also tended to underrepresent ideology 2. It instead overrepresented ideology 2, electing 3 candidates of ideology 2 in 50\% of elections. This indicates that the Schulze method is not a representative method in a multi-seat election. The STV result showed less misrepresentation than the FPTP result, which was to be expected since the method has access to more data of any voter's preferences. The results in scenario 2 can be observed in figure~\ref{fig:scenario 2 result} and the key metrics can be seen in table~\ref{tab:scenario 2 result}.
\begin{figure}
	\centering
	\begin{tikzpicture}
		\begin{axis}[
			ybar,
			width = 0.9\textwidth,
			height = 0.5\textwidth,
			legend style={at={(0.5,-0.25)},
			anchor=north,legend columns=-1},
      x tick style = transparent,
			ylabel = {Elections won},
			xlabel = {Candidate},
			ymajorgrids = true,
			every axis plot/.append style={draw=none,fill,no markers},
      ymin = 0,
			symbolic x coords={0,1,2,3,4,5,6,7,8},
			bar width=0.25cm
      ]
			\addplot coordinates
			{(0, 973)(1, 267)(2, 1000)(3, 0)(4, 12)(5, 743)(6, 0)(7, 5)(8, 0)};
			\addplot coordinates
			{(0, 456)(1, 808)(2, 1000)(3, 0)(4, 33)(5, 703)(6, 0)(7, 0)(8, 0)};
			\addplot coordinates
			{(0, 338)(1, 160)(2, 1000)(3, 0)(4, 0)(5, 1000)(6, 0)(7, 502)(8, 0)};
			\legend{FPTP, STV, Schulze}
		\end{axis}
	\end{tikzpicture}
	\caption{Results in scenario 2}
\label{fig:scenario 2 result}
\end{figure}

\begin{table}
\centering
\caption{Key metrics for result in scenario 2}
\label{tab:scenario 2 result}
\begin{tabular}{@{}lccc@{}}
\toprule
Method & Standard deviation & Misrepresentation & Most likely winners \\ \midrule
FPTP & 418 & 50 & 0, 2 \& 5 \\
STV & 388 & 47 & 1, 2 \& 5 \\
Schulze & 393 & 100 & 2, 5 \& 7  \\ \bottomrule
\end{tabular}
\end{table}
% [ 418.13628234977, 388.17206952015, 393.25535950293 ]
\subsection{Scenario 3}
Scenario 3 featured 9 candidates of which 3 were to be elected. The distribution of votes was steep. This can be seen by the first-preference votes in figure~\ref{fig:example of scenario 3}. The results obtained with each method in this scenario are practically identical. From this, it can be inferred that the results of the methods converge as the preferences in a population become more clear. The results obtained in this scenario can be seen in figure~\ref{fig:scenario 3 result}. The key metrics can be found in table~\ref{tab:scenario 3 result}.
\begin{figure}[H]
	\centering
	\begin{tikzpicture}
		\begin{axis}[
			ybar,
			width = 0.9\textwidth,
			height = 0.5\textwidth,
			legend style={at={(0.5,-0.25)},
			anchor=north,legend columns=-1},
      x tick style = transparent,
			ylabel = {Elections won},
			xlabel = {Candidate},
			ymajorgrids = true,
			every axis plot/.append style={draw=none,fill,no markers},
      ymin = 0,
			symbolic x coords={0,1,2,3,4,5,6,7,8},
			bar width=0.25cm
      ]
			\addplot coordinates
			{(0, 994)(1, 1000)(2, 1000)(3, 0)(4, 0)(5, 6)(6, 0)(7, 0)(8, 0)};
			\addplot coordinates
			{(0, 1000)(1, 1000)(2, 1000)(3, 0)(4, 0)(5, 0)(6, 0)(7, 0)(8, 0)};
			\addplot coordinates
			{(0, 999)(1, 1000)(2, 1000)(3, 0)(4, 0)(5, 1)(6, 0)(7, 0)(8, 0)};
			\legend{FPTP, STV, Schulze}
			\end{axis}
	\end{tikzpicture}
	\caption{Results in scenario 3}
\label{fig:scenario 3 result}
\end{figure}

\begin{table}[H]
	\centering
	\caption{Key metrics for result in scenario 3}
\label{tab:scenario 3 result}
	\begin{tabular}{@{}lccc@{}}
		\toprule
		Method & Standard deviation & Misrepresentation & Most likely winners \\ \midrule
		FPTP & 470 & 0 & 0, 1 \& 2 \\
		STV & 471 & 0 & 0, 1 \& 2 \\
		Schulze & 471 & 0 & 0, 1 \& 2  \\ \bottomrule
	\end{tabular}
\end{table}
\section{Conclusion}
The results obtained in the three different scenarios provide some insight into the varying properties of the different voting systems. Each of the scenarios showed different behaviors for the different methods. For instance, the methods provided practically the same result in scenario 3 but wildly varying results in scenario 2. While the testing method used in this paper may not have been the most robust way of comparing the voting methods, several conclusions can be drawn from the data. I outline them in 4 points below.
\begin{enumerate}
	\item Schulze did not seem like a proportional voting method in multi-seat elections. ...
	\item In single-seat elections, Schulze and STV (and IRV in extension) seemed to provide similar results. ...
	\item FPTP did not seem like a very proportional method in close elections. ...
	\item STV seems to be the most proportional method out of the 3 tested, but is not without its flaws. ...
\end{enumerate}
Summary:
There is no perfect voting system. Generally, if you want proportional representation (which in a democracy you want), STV will likely be the best method to use out of the methods covered in this paper.
Modeling of data needs more work, as well as including other methods (such as \ldots). More research is needed.

Todo:
\begin{itemize}
	\item Write theory
	\item Add references where needed
	\item Explain metrics used to evaluate results
	\item Cut down on explanation of fptp
	\item Write abstract
	\item Finish summary
\end{itemize}

Hello world

>>>>

\begin{abstract}
\end{abstract}

@@

\section{Introduction}
\subsection{Background}
Taking a collective decision as a population is difficult. To solve this issue, voting systems with defined rules are used. They are used to show common preferences within a population, for example what politician a population wants to see elected. Several types of systems have been designed for this purpose and there are a myriad of variations of those systems. They range from simple methods such as “most votes win” to complex processes that can only be practically carried out by a computer. However, practically all voting systems are algorithmic in nature, which makes them interesting to study for both Computer Scientists and Mathematicians. When deciding on what systems that shall be used, it is useful to know what differences and similarities they have.
\subsection{Purpose}
The purpose of this essay is to evaluate differences and similarities in terms of election results in three different voting systems: Single Transferable Vote (STV), First-Past-The-Post (FPTP) and the Schulze Method. This paper will only be examining and discussing differences and similarities in the results of the methods, not touching on how practically feasible the methods might be in actual election.
\pagebreak
\section{Theory}
One would think that by testing all different voting systems and by setting up good criteria for how to evaluate them, it would be possible to come to a definitive conclusion of what system should be used. However, this is not the case. By setting up different mathematical criteria for how a voting system shall behave, it is impossible to satisfy all criteria simultaneously. This is what several theorems, most notably \textit{Arrow's Impossibility Theorem} show \autocite{arrow1950difficulty}. This means that any decision regarding choice of electoral system is a balance of their strengths and weaknesses. Simply put: there is no perfect voting system and never will be. This does however not mean that all voting systems are equally efficient, elegant or accurate.

There are several ways of evaluating voting methods that are used. One is the mathematical approach of evaluating what criteria are important to meet and finding methods that satisfy them. Another is by testing and finding probabilities that some criteria are broken — A method that fails to meet a certain criteria only in very few cases is perhaps not that big of a problem. A third way of evaluating methods is not by mathematical criteria, but other metrics such as voter satisfaction or how likely someone is to vote tactically.

@@

All voting algorithms have been implemented in Typescript – mostly due to the author having previous experience with JavaScript. Typescript is a superset of JavaScript with multiple extra features. Most importantly Typescript has optional static type checking. Sample code written in Typescript can be seen in code example~\ref{lst:typescript example}.

@@

\label{sec:vote-generation}
In order to test the methods and their implementations, test data is needed. There are cases where extensive election data is published, such as in Maltese elections, but full ballot data, which is needed as input for the implementations used in this paper has not been found. Instead, the algorithms will be tested with data generated from a simple computer program. The algorithm used for this purpose in this paper is highly primitive, as modeling preferences within a population is far beyond the scope of this paper. The full ballot generation program can be found in the \Code{/generator} folder. The goal of the program is to produce ballots that rank every candidate in order of preference (see figure~\ref{fig:stv ballot} on page~\pageref{fig:stv ballot}). The program incorporates the idea of an \textit{ideology} where each candidate is assigned one. Each ideology has has a list of weighted preferences indicating how a voter for this ideology is likely to vote. The weighed preferences are generated via an abstraction. A single configuration object is used to create all ideologies.

@@

	\item Based on the base size of each candidate, a probability-list of voting for each candidate is created for every ideology. An ideology's probability-list represents how a voter aligned the ideology is likely to vote.
	\item It creates a certain number of individual ballots. Each ballot is given an ideology and creates its own list preferences via weighted random number generation according to the probability-list of its assigned ideology.

@@

The output of the program is a set of ballots, where each ballot is an ordered set of preferences between candidates. The implementation is highly arbitrary and all constants in the program even more so. However, expanding on the generator program would be beyond the scope of this project. The vote generation algorithm is an attempt to provide more realistic voting data than what is achieved by a purely random allocation of votes. The program also allows for the possibility to control the data by changing the input variables. This makes it possible to test different election scenarios.

@@

Constructing test scenarios is difficult task. Due to there being so many variables both in real elections and the simulated elections in the generator program, it makes it practically impossible to control all variables. Instead, 3 arbitrary test scenarios will be used. An election will be simulated 1000 times for each scenario and the election data will then run through the different voting algorithms and the results will be compared. The simulation scenarios are arbitrary but should allow for comparison of the different voting methods. All simulations will have 3 ideologies and 500 voters.

@@

The scenario is built around electing one candidate out of eight. The spread of votes within ideologies is flat and all ideologies are so called ``weak''. This means that the general spread of votes between candidates will be even. An example of the distribution of first-preference votes generated in this scenario can be seen in figure~\ref{fig:example of scenario 1}. The full configuration object can be seen in code example~\ref{lst:scenario 1}.

@@

	\caption{Example distribution of first-preference votes in scenario 1. Avreages of 10 simulations.}

@@

	\caption{Example distribution of first-preference votes in scenario 2. Avreages of 10 simulations.}

@@

	\caption{Example distribution of first-preference votes in scenario 3. Avreages of 10 simulations.}

@@

Three metrics will be used for comparing the results obtained in the different scenarios. The results obtained with each voting method will be compared against the other methods. Comparisons to the configuration files outlined in Sections~\ref{sec:method-scenario1} -~\ref{sec:method-scenario3} will also be made.

@@

The standard deviation of the outcomes represents the variance in outcome due to the pseudorandom nature of the generator program outlined in Section~\ref{sec:vote-generation}. The standard deviation is calculated according to equation~\ref{eq:standard-deviation}. 

@@

	\item Compare the actual outcomes of a certain method with the theoretical result obtained in Step~\ref{enum:theoretical}.

@@

The first-past-the-post voting method is a method designed for electing one or several candidates out of a set of candidates. Each voter has one (1) vote that may be allocated to any candidate. In a FPTP election the N candidates with most votes get elected.

@@

The first-past-the-post program is implemented in a single file (\Code{/fptp/index.ts}). The inputs of the program can be seen in code example~\ref{lst:fptp inputs}.

@@

The Single Transferable Vote (STV) is a more complicated voting system than the first-past-the-post method. STV is a multi-seat election method, although it can be used for single-seat elections too. STV becomes equivalent to the Instant-runoff voting method (IRV) in single-seat elections. In an STV election, voters are given ballots in which they are supposed to rank candidates in order. The specific rules for wether one needs to rank all candidates or just a few not can vary. For the sake of simplicity, all candidates must be ranked at unique and linear values in this implementation. An example of an STV ballot can be seen in figure~\ref{fig:stv ballot}. By having voters convey more information in a ballot, there is more data available that the method can take advantage of. 

@@

Using this data, votes are transferred between candidates within a ballot in the method, hence the name Single Transferable Vote. The method relies on the concept of a quota, which is the number of votes you need to be elected. The Droop-Quota, defined as $(\frac{\text{total valid poll}}{\text{seats} + 1})+1$, is often used and is the one implemented in this paper. If a candidate receives equal or more votes than the quota, he/she is elected. If the number of votes for a candidate exceed the quota, a fraction of his/her votes are transferred to the next choice on the ballots that voted for that particular candidate. This process is repeated until there are no candidates with more votes than the quota. If there still are seats yet to be filled, the candidate with fewest votes is eliminated and his/her votes are transferred to the next choice on the ballots which supported the candidate. This process is repeated until all seats are filled. The process can become quite complex with many ballots and transferring of fractional votes. A computer is therefore essential to resolve large scale elections.

@@

\tab eliminate $votes_{k}$ and distribute all next-preference votes for candidate \\
\tab $k$ into $votes$\\

@@

The single transferable vote algorithm uses a tree structure to represent the state of the election. Every candidate is a top-level node with an attribute showing the number of current votes for the candidate. Each node has child nodes representing the next preferences of the voters for that particular candidate, see figure~\ref{fig:tree structure}. It is easy to manipulate branches via recursive functions. Merging and multiplying branches is used when transferring votes from one candidate node to another. A candidate node is a Typescript class. It can be found in the \Code{stv/candidate.ts} file and has the following properties and functions:

@@

The Schulze Method is a method used mainly to determine results in single-winner elections but can also be used to provide rankings and elect multiple candidates in a single election \autocite{Schulze2011}. Ballots are identical to those of the single transferable vote (see figure~\ref{fig:stv ballot} on page~\pageref{fig:stv ballot}). The method is compliant with the Condorcet Criterion, which means that if there is a candidate who the majority prefers in a pairwise comparison with every other candidate, that candidate wins. As mentioned, this method relies on comparing the preferences of every candidate to one another as shown in table~\ref{tab:pairwise comparison matrix}. If there is a Condorcet winner the process is simple: declare the Condorcet winner a winner and run another iteration without that candidate. Repeat this process until all seats are filled. However, there can be Condorcet ties, where there is no Condorcet winner. There are multiple methods used for resolving Condorcet ties, with one of them being the Schulze Method.

@@

The difficult problem in this method is finding the strongest path between candidates. The problem is called the widest path problem in graph theory. An efficient and relatively simple way to compute this problem is via the Floyd-Warshall algorithm which is used in the implementation. See Code Example~\ref{lst:schulze algorithm} for the full algorithm.

@@

let $C$ be the number of candidates

@@

The Schulze Method is implemented within the file \Code{schulze/index.ts}. It takes two inputs: a list of ballots and the number of seats to be elected. A ballot is an ordered list of candidates. The output of the program is a list of winners. The program first creates the matrices for both the preferences and paths between candidates as seen in Code Example~\ref{lst:schulze-mapsb}.
\codeblock{code/schulze/maps.ts}{lst:schulze-maps}{Data representation}

@@

\label{sec:resuls-scenario1}
Scenario 1 featured 8 candidates where one was to be elected. Vote spread was flat, meaning that votes tend to be diverse in their preferences. This means that there is not an obvious winner present. This can be seen from the first-preference votes in the scenario seen in figure~\ref{fig:example of scenario 1} on Page~\pageref{fig:example of scenario 1}. The full result in this scenario can be seen in figure~\ref{fig:scenario 1 results}. The outlier in this scenario seems to be the FPTP method, electing candidate 0 more often than both other methods. Key metrics for the results in this scenario can be found in table~\ref{tab:scenario 1 result}. Although the misrepresentation metric is not entirely accurate, especially when few seats are elected, it shows that fptp is not entirely representative. The STV and Schulze method seem to produce very similar results.

@@

\begin{table}[H]
\centering
\caption{Key metrics for result in scenario 1}
\label{tab:scenario 1 result}
\begin{tabular}{@{}lccc@{}}
\toprule
Method & Standard deviation & Misrepresentation & Most likely winners \\ \midrule
FPTP & 223 & 66 & 2 \\
STV & 315 & 8 & 2 \\
Schulze & 313 & 9 & 2 \\ \bottomrule
\end{tabular}
\end{table}
% [ 223.39259164082, 315.44928276983, 313.6271671906]
\subsection{Scenario 2}
\label{sec:resuls-scenario2}
Scenario 2 featured 9 candidates of which 3 were to be elected. The distribution of votes was ``normal'', and as seen in figure~\ref{fig:example of scenario 2} on page~\ref{fig:example of scenario 2}, the election is very close with many candidates having a similar percentage of first-preference votes. The results obtained in scenario 2 are very interesting. FPTP tended to underrepresent candidate 1, and elected a candidate from ideology 2 only in 28\% of elections, despite the ideology theoretically having 30\% of popular support. The Schulze method also tended to underrepresent ideology 2. It instead overrepresented ideology 3, electing 3 candidates of ideology 3 in 50\% of elections. This indicates that the Schulze method is not a representative method in a multi-seat election. The STV result showed less misrepresentation than the FPTP result, which was to be expected since the method has access to more data of any voter's preferences. The results in scenario 2 can be observed in figure~\ref{fig:scenario 2 result} and the key metrics can be seen in table~\ref{tab:scenario 2 result}.
\begin{figure}[H]
	\centering
	\begin{tikzpicture}
		\begin{axis}[
			ybar,
			width = 0.9\textwidth,
			height = 0.5\textwidth,
			legend style={at={(0.5,-0.25)},
			anchor=north,legend columns=-1},
      x tick style = transparent,
			ylabel = {Elections won},
			xlabel = {Candidate},
			ymajorgrids = true,
			every axis plot/.append style={draw=none,fill,no markers},
      ymin = 0,
			symbolic x coords={0,1,2,3,4,5,6,7,8},
			bar width=0.25cm
      ]
			\addplot coordinates
			{(0, 973)(1, 267)(2, 1000)(3, 0)(4, 12)(5, 743)(6, 0)(7, 5)(8, 0)};
			\addplot coordinates
			{(0, 456)(1, 808)(2, 1000)(3, 0)(4, 33)(5, 703)(6, 0)(7, 0)(8, 0)};
			\addplot coordinates
			{(0, 338)(1, 160)(2, 1000)(3, 0)(4, 0)(5, 1000)(6, 0)(7, 502)(8, 0)};
			\legend{FPTP, STV, Schulze}
		\end{axis}
	\end{tikzpicture}
	\caption{Results in scenario 2}
\label{fig:scenario 2 result}
\end{figure}

\begin{table}[H]
\centering
\caption{Key metrics for result in scenario 2}
\label{tab:scenario 2 result}
\begin{tabular}{@{}lccc@{}}
\toprule
Method & Standard deviation & Misrepresentation & Most likely winners \\ \midrule
FPTP & 418 & 50 & 0, 2 \& 5 \\
STV & 388 & 47 & 1, 2 \& 5 \\
Schulze & 393 & 100 & 2, 5 \& 7  \\ \bottomrule
\end{tabular}
\end{table}
% [ 418.13628234977, 388.17206952015, 393.25535950293 ]
\subsection{Scenario 3}
\label{sec:resuls-scenario3}
Scenario 3 featured 9 candidates of which 3 were to be elected. The distribution of votes was  ``steep''. This can be seen by the first-preference votes in figure~\ref{fig:example of scenario 3} on Page~\pageref{fig:example of scenario 3}. The results obtained with each method in this scenario are practically identical. From this, it can be inferred that the results of the methods converge as the preferences in a population become more clear. The results obtained in this scenario can be seen in figure~\ref{fig:scenario 3 result}. The key metrics can be found in table~\ref{tab:scenario 3 result}.
\begin{figure}[H]
	\centering
	\begin{tikzpicture}
		\begin{axis}[
			ybar,
			width = 0.9\textwidth,
			height = 0.5\textwidth,
			legend style={at={(0.5,-0.25)},
			anchor=north,legend columns=-1},
      x tick style = transparent,
			ylabel = {Elections won},
			xlabel = {Candidate},
			ymajorgrids = true,
			every axis plot/.append style={draw=none,fill,no markers},
      ymin = 0,
			symbolic x coords={0,1,2,3,4,5,6,7,8},
			bar width=0.25cm
      ]
			\addplot coordinates
			{(0, 994)(1, 1000)(2, 1000)(3, 0)(4, 0)(5, 6)(6, 0)(7, 0)(8, 0)};
			\addplot coordinates
			{(0, 1000)(1, 1000)(2, 1000)(3, 0)(4, 0)(5, 0)(6, 0)(7, 0)(8, 0)};
			\addplot coordinates
			{(0, 999)(1, 1000)(2, 1000)(3, 0)(4, 0)(5, 1)(6, 0)(7, 0)(8, 0)};
			\legend{FPTP, STV, Schulze}
			\end{axis}
	\end{tikzpicture}
	\caption{Results in scenario 3}
\label{fig:scenario 3 result}
\end{figure}

\begin{table}[H]
	\centering
	\caption{Key metrics for result in scenario 3}
\label{tab:scenario 3 result}
	\begin{tabular}{@{}lccc@{}}
		\toprule
		Method & Standard deviation & Misrepresentation & Most likely winners \\ \midrule
		FPTP & 470 & 0 & 0, 1 \& 2 \\
		STV & 471 & 0 & 0, 1 \& 2 \\
		Schulze & 471 & 0 & 0, 1 \& 2  \\ \bottomrule
	\end{tabular}
\end{table}
\section{Conclusion}
The results obtained in the three different scenarios provide some insight into the varying properties of the different voting systems. Each of the methods showed different behavior in the different scenarios. For instance, the methods provided practically the same result in scenario 3 but wildly varying results in scenario 2. While the testing method used in this paper may not have been the most robust way of comparing the voting methods, several conclusions can be drawn from the data. They are outlined in 4 points below.
\begin{enumerate}
	\item Schulze did not seem like a proportional voting method in multi-seat elections.\label{num:conclusion-schulze}

	As seen from the results obtained in scenario 2, which can be seen in Section~\ref{sec:resuls-scenario2}, the Schulze Method heavily overrepresented candidates from ideology 1 (candidates 3, 5 \& 7). While ideology 1 in fact was the most popular ideology, a more proportionate way of representing the electorate in the scenario would have been to grant one seat to each ideology. The Schulze tended to elect candidates from the strongest ideology only, instead of electing candidates from all ideologies in a more proportionate manner. This is not a problem in a single-seat election, as can be seen in the results of scenario 1 in Section~\ref{sec:resuls-scenario1}, which was a single-seat elections, where the Schulze Method simply elected the most popular candidate.
	\item In single-seat elections, Schulze and STV (and IRV in extension) seemed to provide similar results.

	As seen in the results of scenario 1 outlined in Section~\ref{sec:resuls-scenario1}, both Schulze and STV produced nearly identical results, while the results from FPTP deviated slightly. While the methods work very differently, the results in these methods seem to behave similarly in single-seat elections.
	\item FPTP did not seem like a very proportional method in close elections.

	FPTP preformed considerably worse at being representative than the other methods. This was true for both close single- and multi-seat elections, excluding The Schulze Method in scenario 2 (see Point~\ref{num:conclusion-schulze}). At the core, both the STV and Schulze Method have more data to use for determining winners than what FPTP does. The FPTP method only has access to first-preference votes which does not convey as much information as a ballot listing all candidates in order of preference does.
	\item STV seems to be the most proportional method out of the 3 tested, but is not without its flaws.

	STV provided the most proportionate results overall of all three methods tested. However, as it goes with voting system, it has several flaws. It is for example possible to increase the chance of a candidate winning by lowering him/her on a ballot, and vice versa. STV is the only multi-seat proportional voting system tested in this paper, and the results obtained reinforce its position as just that.
\end{enumerate}
While there is no perfect voting system, they all preform differently. From the conclusions outlined above, a decision of what system to use between the three voting systems used in this paper can be taken. Generally, if a proportional representation is desired (which in a democracy it usually is), STV will likely be the most appropriate method to use. In single-seat elections, the Schulze Method should also be considered. FPTP is the simplest method to both use and understand, but cannot be considered a proportionate voting method.

While the findings of this paper are hardly revolutionary, the nature of the three voting systems used in this paper and their behavior is experimentally evaluated. The paper also provides implementations of all voting methods in both TypeScript and JavaScript.

The key area for furthering this work would be in data collection or creation. A more sophisticated procedure for generating test data could be developed and run with more extensive test scenarios. Data from the real world could also be collected in actual elections and run through the different algorithms to test them. One could also write software that generates lots of scenarios with different parameters and analyzes how they differ automatically. Since the source code for this paper is open source under the MIT license, anyone can freely copy, modify and build on this work. This also makes for an open and transparent voting method where every voter can audit the system in use.
\pagebreak
